Creating Dynamic Documents
======================================================================================
Embedding Code Chunks in Scientific Documents
----------------------------------------------------------

Chris Paciorek, Department of Statistics, UC Berkeley

```{r setup, include=FALSE}
opts_chunk$set(engine='bash') # because we're using a lot of bash, let's set as default
```

# 0) This Tutorial

This tutorial covers the basics of creating documents that combine code chunks, mathematical notation, and text. We'll cover R and bash shell chunks (plus a little bit on Python) in the context of documents written with Latex and Markdown. 

A screencast of the material accompanies this document, and the end of this document has questions that test your understanding of the material. 

We'll use a virtual machine developed here at Berkeley, [the Berkeley Common Environment (BCE)](http://bce.berkeley.edu). BCE is a virtual Linux machine - basically it is a Linux computer that you can run within your own computer, regardless of whether you are using Windows, Mac, or Linux. This provides a common environment so that things behave the same for all of us.

Materials for this tutorial, including the R markdown file that was used to create this document are available on github at (https://github.com/berkeley-scf/tutorial-dynamic-docs).  You can download the files by doing a git clone from a terminal window on UNIX-like machine, as we'll demonstrate early in this document:
```{r, clone, eval=FALSE}
git clone https://github.com/berkeley-scf/tutorial-dynamic-docs
```

This is a pull-yourself-up-by-your-bootstraps tutorial as this document itself is generated from an R Markdown file in the same way as we discuss herein.

To create this HTML document, simply compile the corresponding R Markdown file in R as follows (the following will work from within BCE after cloning the repository as above).
```{r, build-html, eval = FALSE}
Rscript -e "library(knitr); knit2html('dynamic-docs.Rmd')"
```

<!--
#pandoc --number-sections spark.md -o spark.html
#Rscript -e "library(knitr); knit('spark.Rmd')"
#pandoc --mathjax --number-sections spark.md -o spark.html
-->



# 1) Getting prepared

First, you'll need to get the BCE virtual machine running on your computer. First install the VirtualBox software on your local machine. BCE runs on VirtualBox. Please follow the [BCE installation instructions](http://bce.berkeley.edu/install). Once you've done this, you'll have a Linux computer running within your own machine.

That said, if you'd like to work on your own machine, you can. But you'll need to install the following:
* a LaTeX installation
* R (and optionally RStudio)
* the knitr package for R
% * Jupyter and the R kernel for Jupyter 

Note that more details on LaTeX are available in our [https://github.com/berkeley-scf/tutorial-latex-intro](quick introduction to LaTeX).

# 5) RMarkdown: Markdown + R

# 1) Latex + R using knitr (Rtex format)

# adding bash or Python chunks

# 1) Latex + R using Sweave formatting (Rnw format)

# adding bash or Python chunks

# 4) Lyx + R

